### 2023-12-26 ms.qmd
Updates

Cognition is defined as the processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision making (). These mechanisms underpin several aspects of animals' ecology such as habitat selection, foraging decisions, mate choice, antipredatory strategies, and/or social behaviours that are crucial for survival and reproduction (@dukas_evolutionary_2004). Particularly, learning - the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) - is seen as fundamental for coping with environmental changes by allowing individuals to create new associations between events (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013). However, the ability of organisms to acquire new information varies accross individuals, with this natural variation usually being influenced by factors like age, sex, social system, gut microbiota, or the environment where animals develop (**REFS**). In this regard, the latter can be particularly relevant because the brain is especially sensitive to environmental impacts during early stages of development (@zhu_prenatal_2004). 
These differences in neural development caused by early environment can have long-lasting effects on learning skills, and prenatal Glucocorticoids (GCs) and prenatal thermal environment are known to play a prominent role in shaping learning abilities in different taxa (see @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @crino_corticosterone_2014; @abayarathna_effects_2020). GCs - **hormones related to organisms' response to changes in homeostasis (BEST DEF AND REF**) - can have sustained pysiological effects on neural structure and physiology associated with animals' performance on learning tasks (**Definition and REF**). In this sense, there is evidence that high prenatal GC levels impair with learning accross different taxa (**MORE EG.S AND REFS**; @lemaire_prenatal_2000; @zhu_prenatal_2004), but results vary depending on sex, or task (**REFS**). Also, some studies have revealed effects of prenatal temperature on learning in ectotherms (**REFS**). For instance, ... (**REFS**). In this vein, the effect of prenatal temperature has been associated with similar changes in neural structure and metabolic activity (**REFS**). However, despite both prenatal GCs and thermal environment are known to affect learning through similar mechanisms, the interaction between this two factors is still poorly understood.
In this study, we aim to explore the interactive effects between prenatal Glucocorticoids (GCs) and the prenatal thermal environment on learning using two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_), as our model species. We experimentally increased Corticosterone (CORT) - the main GC in reptiles - levels in the eggs of these two species of skinks and then incubated them at two different temperatures in a 2X2 factorial design. Then, we subjected the juveniles to a colour-associative and a reversal task to explore their learning abilities. We hypothesise that changes in CORT levels and temperature during early development will have sustained effects on brain's morphology and physiology that will ultimately impact learning skills. We predicted that individuals exposed to high levels of CORT and/or low temperatures will perform worse in the learning tasks than control individuals or individuals exposed to high temperatures. We also expected incubation at high temperatures to decrease the impact of CORT on skink's performance, while cold incubation temperatures were expected to increase CORT detrimental effects on learning.

**Cognition it is affected by the conditions in which organisms develop (Burger, 1991; Lemaire et al., 2000; Amiel and Shine, 2012; Amiel et al., 2014; Abayarathna and Webb, 2020).**

@@

Two weeks before we started the training phase (see below), lizards were moved to the experimental arena (Fig. 1c) for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the other. These new enclosures were placed in two rooms in 7 different racks associated to 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and all the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied *ad libitum.* We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The rooms temperature was set to between 22-24 Celsius. We tested temperatures with laser temperature guns to ensure such a set up established a thermal gradient and we there were no significant differences between treatments or species (see Supplementary Material).

@@

To estimate learning skills, we tested skinks' ability to locate a food reward in a series of behavioural tasks (Fig. 2) (see [@leal_behavioural_2012]; [@clark_colour_2014]). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical to the ones from the experiment except for the colour (see below). We divided this training phase into three stages (see Fig. 2): in the first stage, lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure; in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps; and finally, the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp in the third and last stage. Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of the experiment, we recorded whether the cricket has been consumed or not. Trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials. This phase lasted 38 days until all the lizards were able to eat from the ramp; only in one case we decided not to use the lizard because its behaviour was not consistent over the course of the training phase (see Supplementary material).

In the next phase, we trained lizards to associate between colour and a food reward (colour association task in Fig. 2). The test was like the third stage of the training phase, but here lizards were presented with three feeders that differed in the colour. We placed the food reward (small, frozen, *A. domestica* crickets) inside the wells of the three feeders, covering two of the crickets with 3D-printed lids (3D x 0.5H cm) so prey was only accessible in one ramp. The food reward was placed in all three wells to avoid the use of chemical cues to determine the correct choice; our lids had a series of small holes on the top for the same reason. The colours of the feeders were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours (REF). To control for potential colour preference that could bias our results, we split the subjects in two groups counterbalanced by treatment and species; in one group the correct choice (i.e., the ramp with non-covered cricket) was the blue one in the associative task and red in the reversal, while we did the opposite for the other half. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in this task once a day for 35 days.

@@

INSERT Fig 2

>>>>

Cognition is defined as the processes by which animals gather, preserve, and use information from their environment through perception, learning, memory, and decision making (@shettleworth). These mechanisms underpin several aspects of animals' ecology such as foraging, mate choice, antipredatory strategies, and/or social behaviours that are crucial for survival and reproduction (@dukas_evolutionary_2004). Particularly, learning - the acquisition of neuronal representations of new information (@dukas_evolutionary_2004) - is seen as fundamental for coping with environmental changes by allowing individuals to create new associations between events (@dukas_evolutionary_2004; @leal_behavioural_2012; @buchanan_condition_2013). However, the ability of organisms to acquire new information varies accross individuals, with this natural variation usually being influenced by factors like age, sex, gut microbiota, or the environment where animals develop (@lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014). In this regard, the latter can be particularly relevant because the brain is especially sensitive to environmental impacts during early stages of development (@zhu_prenatal_2004). 
Prenatal Glucocorticoids (GCs) and prenatal thermal environment are known to play a prominent role in shaping learning abilities in different taxa (see @lemaire_prenatal_2000; @zhu_prenatal_2004; @amiel_hotter_2012; @amiel_egg_2014; @crino_corticosterone_2014; @abayarathna_effects_2020). GCs - hormones related to organisms' response to stress (@sapolsky_how_2000) - can have sustained effects on neural structure and physiology that are associated with animals' performance on learning tasks (@lemaire_prenatal_2000; @zhu_prenatal_2004; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). Some studies have demonstrated that prenatal stress and high prenatal GC levels impair with learning (@lemaire_prenatal_2000; @zhu_prenatal_2004; @farrell_developmental_2015-learn), while others revealed that these effects may vary depending on aspects like sex or learning task (@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016). Similarly, diverse experiments have shown pivotal impacts of prenatal temperature on learning in ectotherms (@amiel_hotter_2012; @amiel_egg_2014;). For instance, high incubation temperatures have been linked with faster learning rates in skinks (@amiel_hotter_2012; @amiel_egg_2014), but velvet geckos incubated at temperatures over their natural range learn slower than those incubated within the natural thermal limits (@abayarathna_effects_2020). In this vein, the effect of prenatal temperature has been linked to alterations in neural structure and metabolic activity (@coomber_independent_1997; @sakata_neural_2000; @amiel_effects_2017; @beltran_are_2021) that are similar to those resulting from prenatal increased stress or GC levels (@lemaire_prenatal_2000; @zhu_prenatal_2004; @du_dynamic_2009). This suggests that prenatal GCs and temperature can act on the same physiological mechanisms and, as such, both may interact to shape individual variation on learning skills (@amiel_effects_2017; @noble_developmental_2018). Furthermore, GCs can play an important role in determining vertebrate responses to elevated temperatures (@Crino_2023) potentially driving to natural interactions between temperature and GCs. However, despite the similar mechanistic effects of prenatal GCs and temperature and the potential function of GCs in vertebrates response to elevated temperatures, our understanding of how these two factors interplay remains incomplete.
In this study, we aim to explore the interactive effects between prenatal Glucocorticoids (GCs) and the prenatal thermal environment on learning using two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_), as our model species. We experimentally increased Corticosterone (CORT) - the main GC in birds, reptiles, amphibians, and rodents (@Crino_2023) - levels in the eggs of these two species of skinks and then incubated them at two different temperatures in a 2X2 factorial design. Then, we subjected the juveniles to a colour-associative and a reversal task to explore their learning abilities. We hypothesised that changes in CORT levels and temperature during early development would have sustained effects on brain's morphology and physiology that would ultimately impact learning skills. We predicted that individuals exposed to high levels of CORT and/or low temperatures would perform worse in the learning tasks than control individuals or individuals exposed to high temperatures. We also expected incubation at high temperatures to decrease the impact of CORT on skink's performance, while cold incubation temperatures are expected to increase CORT detrimental effects on learning.

@@

Two weeks before we started the training phase (see below), lizards were moved to the experimental arena  for acclimatation. The arenas were individual medium size (41 L x 29.7 W x 22 H cm) plastic containers with a shelter (9 L x 6 W x 1.5 H cm) on one of the extremes and a water dish on the other. These new enclosures were placed in two rooms in 7 different racks associated to 7 different CCTV systems (device model DVR-HP210475) that allowed us to record their behaviour during the experiment (see details below). The number of lizards per species and treatment in each rack was counterbalanced to control for any effect of the room or the position of the lizard in the rack. During acclimatation and all the experiment, lizards were fed with only one cricket per day dusted with calcium and multivitamin (see protocol below), and water was supplied *ad libitum.* We provided a temperature gradient by means of a heat cord and heat lamps in a 12 h light: 12 h dark cycle. The rooms temperature was set to between 22-24 Celsius. We tested temperatures with laser temperature guns to ensure such a set up established a thermal gradient and we there were no significant differences between treatments or species (see Supplementary Material).

@@

To estimate learning skills, we tested skinks' ability to locate a food reward in a series of behavioural tasks (Fig. 1b) (see [@leal_behavioural_2012]; [@clark_colour_2014]). First, we performed a training phase where lizards had to learn to eat from white 3D-printed PLA ramps (9 L x 4 W x 5 H cm) identical to the ones from the experiment except for the colour (see below). We divided this training phase into three stages (see Fig. 1b): in the first stage, lizards had to eat a small, frozen cricket (*A. domestica*) from an opaque petri dish (3D x 1.6H cm) placed in the middle of their enclosure; in the second stage, the petri dish with the cricket was placed on top of the white 3D printed ramps; and finally, the cricket was left inside a well (3D x 1.75H cm) on the top of the ramp in the third and last stage. Every trial began when we left the feeding block (petri dish, ramp, or both) inside the enclosure and finished one hour later when we took it away. At the end of the experiment, we recorded whether the cricket has been consumed or not. Trial was considered successful if the lizard could locate and consume the reward, while completion of each stage required the lizards to eat the crickets in at least 5 out of 6 trials. This phase lasted 38 days until all the lizards were able to eat from the ramp; only in one case we decided not to use the lizard because its behaviour was not consistent over the course of the training phase (see Supplementary material).

In the next phase, we trained lizards to associate between colour and a food reward (colour association task in Fig. 1b). The test was like the third stage of the training phase, but here lizards were presented with three feeders that differed in the colour. We placed the food reward (small, frozen, *A. domestica* crickets) inside the wells of the three feeders, covering two of the crickets with 3D-printed lids (3D x 0.5H cm) so prey was only accessible in one ramp. The food reward was placed in all three wells to avoid the use of chemical cues to determine the correct choice; our lids had a series of small holes on the top for the same reason. The colours of the feeders were green, red, and blue, as previous studies demonstrate that squamates can discriminate between these colours (REF). To control for potential colour preference that could bias our results, we split the subjects in two groups counterbalanced by treatment and species; in one group the correct choice (i.e., the ramp with non-covered cricket) was the blue one in the associative task and red in the reversal, while we did the opposite for the other half. In all trials, the position of the feeders was changed randomly to ensure subjects were using colour rather than spatial cues for the association. Lizards were tested in this task once a day for 35 days.

@@
